Loan Data From Proper Exploration by NongR
========================================================

```{r, echo=FALSE}
# Workaround for an error when Knit this .rmd file
knitr::opts_chunk$set(error = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)

library(dplyr)
library(lubridate)
library(gridExtra)
library(grid)
library(GGally)
library(RColorBrewer)

```



```{r echo=FALSE, Load_the_Data}
# Load the Data

pl <- read.csv('prosperLoanData.csv')

```



This report explores a dataset containing prices and attributes for 
approximately 114,000 records and 81 variables in this data frame.

```{r echo=FALSE}
dim(pl)
```

Before start analyzing this data, let's update the order of the factor levels of
the variables 'CreditGrade' and create a new variable called 'ListingCategory' 
to store a name of the Listing category

```{r}

# Create a new variable called ListingCategory.
# We will later update this variable to store a name of the Listing Category
pl$ListingCategory <- pl$ListingCategory..numeric.

# Update the Term and ListingCategory columns to be a factor  
pl <- mutate_at(pl, vars(Term, ListingCategory), as.factor)


## --- Update levels
# Update levels of CreditGrade to sorted from Highest to Lowest first
levelsCreditGrade <- c("AA", "A", "B", "C", "D", "E", "HR", "NC", "")

levels(pl$CreditGrade) <- levelsCreditGrade

# Update levels of IncomeRange
levelsIncomeRange <- c("Not displayed", "Not employed", 
                       "$0", "$1-24,999", "$25,000-49,999", 
                       "$50,000-74,999", "$75,000-99,999", "$100,000+")

levels(pl$IncomeRange) <- levelsIncomeRange

# Update levels of ListingCategory

levelsListingCat <- c(
  "Not Available",         #0
  "Debt Consolidation",    #1
  "Home Improvement",      #2
  "Business",              #3
  "Personal Loan",         #4
  "Student Use",           #5
  "Auto",                  #6
  "Other",                 #7
  "Baby&Adoption",         #8
  "Boat",                  #9
  "Cosmetic Procedure",    #10
  "Engagement Ring",       #11
  "Green Loans",           #12
  "Household Expenses",    #13
  "Large Purchases",       #14
  "Medical/Dental",        #15
  "Motorcycle",            #16
  "RV",                    #17
  "Taxes",                 #18
  "Vacation",              #19
  "Wedding Loans")         #20

levels(pl$ListingCategory) <- levelsListingCat

str(pl)
```


# Univariate Plots Section


__CreditGrade__


* Applications with the highest Credit Rating (CreditGrade) which is "AA" has 
the highest record which is expected

```{r echo=FALSE}
ggplot(aes(x=CreditGrade), data=pl) +
  geom_bar()

```


```{r echo=FALSE}

table(pl$CreditGrade)
```


* Interestingly, applications with the medium credit grades ("C", "D", "E")  
have a higher number of loans. I originally expected that this chart should be 
skewed to the right but it is not a case here.

```{r echo=FALSE}
ggplot(aes(x=CreditGrade), data=subset(pl, (CreditGrade!= 'AA' & CreditGrade != ""))) +
  geom_bar()

```

__ListingCreatingDate__

* ListingCreationDate was steady increased during 2006 to early 2008, and 
mid 2009 to late 2012
* (Puzzle) There is a significant drop to almost 0 was seen from mid 2008 to 
mid 2009. Is it because data was not collected during that period?
* We are seeing a spike of number of Listing from 2013 to 2014

```{r echo=FALSE}

grpCreationDate <- pl %>% 
  group_by(date_by_month=floor_date(as.Date(ListingCreationDate), "month")) %>% 
  summarize(n = n())

# Plot chart with plotty
ggplot(aes(x=date_by_month, y=n), data=grpCreationDate) + 
  geom_line() + 
  xlab('Listing Creation by month') +
  ylab('Count') +
  scale_x_date( date_labels = "%Y", date_minor_breaks = "1 month", date_breaks = "1 year")

```

Let's take a closely look at a count of listings of each CreditGrade.

It is very interesting that:

* We only have data of ListingCreationDate for applications with the CreditGrade 
lower than 'AA' from 2006 to 2009
* Data of the CreditGrade 'AA' started from mid 2007 all the way to 2014
* (Puzzle) Is it because of how the system changed at that time or the data 
engineer decided the only collect data with the CreditGrade 'AA' at that point, 
and stopped collecting data for other grades?

```{r echo=FALSE}

grpCreationDate_CreditGrade <- pl %>% 
  group_by(date_by_month=floor_date(as.Date(ListingCreationDate), "month"), 
           CCGrade=CreditGrade) %>% 
  summarize(n = n())

p_grpCreationDate_CreditGrade <- 
  ggplot(aes(x=date_by_month, y=n), data=grpCreationDate_CreditGrade) + 
  geom_line() + 
  xlab('Listing Creation by month') +
  ylab('Count') +
  scale_x_date( date_labels = "%Y",date_breaks = "2 year")

p_grpCreationDate_CreditGrade + facet_wrap(~CCGrade, ncol=3)
```

From the discovery above, I assume that we only need to focus on the CreditGrade 
'AA' from 2010 for this analysis since a number of count before 2010 is 
closed to 0.

See this link for more details about the propser loan

https://www.lendingmemo.com/prosper-loan-complaints/

__New data set__

From this point on, we will analyze data where the ListingCreationDate is 
greater than or equal to 2010 and there are almost 83,000 records in the new 
data set

```{r}

# Get a subset of data which has the ListingCreationData >= 2010
new_pl <- subset(pl, year(ListingCreationDate) >= 2010)
dim(new_pl)
```

__BorrowerRate__

In average, the BorrowRate is around 0.15 except a very high number of 
applications with the Borrow rate around 3.25. Why is that the case?

```{r echo=FALSE}

p1 <- ggplot(aes(x=BorrowerRate), data=new_pl) +
  geom_histogram(binwidth = 0.005) +
  xlim(0, 0.4)

p2 <- ggplot(aes(x='', y=BorrowerRate), 
             data= subset(new_pl, is.finite(BorrowerRate))) +
  geom_boxplot() +
  ylim(0, 0.4) +
  coord_flip()

grid.newpage()
grid.draw(rbind(ggplotGrob(p1), ggplotGrob(p2), size = "last"))

```

```{r echo=FALSE}

summary(new_pl$BorrowerRate)

```


__Occupation__

Let's zoom in data by occupation. The charts below show that most applications
have the Occupations set to 'Other' and the LoanOriginalAmount of this category
is varied a lot comparing to other occupations.

```{r echo=FALSE, fig.width=16, fig.height=12}

grp_Occupation <- new_pl %>% 
  group_by(Occupation) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))

p1 <- ggplot(aes(x=reorder(Occupation, Count), y=Count), data=grp_Occupation) +
  geom_bar(stat = "identity", width=0.5, position=position_dodge(width=0.25)) +
  geom_text(
    aes(label = Count),
    position = position_dodge(0.9),
    hjust = -0.25
  ) +
  ylim(0, 22000) +
  ggtitle('# Applications by Occupation') +
  coord_flip()


p2 <- ggplot(aes(x=reorder(Occupation, Occupation, 
                           function(x) { length(x)}), y=LoanOriginalAmount), 
             data = new_pl) +
  geom_boxplot(width=0.5, position=position_dodge(width=0.25)) +
  ggtitle('LoanOriginalAmount by Occupation') +
  coord_flip()

grid.arrange(p1, p2, ncol=2)
```

__EmploymentStatus__

Most applications have the EmploymentStatus 'Employed' and the median 
LoanOriginalAmount of this category is higher than others.

```{r echo=FALSE, fig.width=16}


p1 <- ggplot(aes(x=reorder(EmploymentStatus, EmploymentStatus, 
                           function(x) { length(x)})), 
             data = new_pl) +
  geom_bar() +
  geom_text(
    stat = 'count',
    aes(label = ..count..),
    position = position_dodge(0.9),
    hjust = -0.25
  ) +
  ylim(0, 70000) +
  ggtitle('# applications by EmploymentStatus') +
  coord_flip()


p2 <- ggplot(aes(x=reorder(EmploymentStatus, EmploymentStatus, 
                           function(x) { length(x)}), 
                 y=LoanOriginalAmount), 
             data=new_pl) +
  geom_boxplot(width=0.5, position=position_dodge(width=0.25)) +
  ggtitle('LoanOriginalAmount by EmploymentStatus') +
  coord_flip()

grid.arrange(p1, p2, ncol=2)
```


__BorrowerState__

Now, let's take a look at a number of Listings by each state. 
Looks like major cities have a higher number of applications.

```{r  echo=FALSE, fig.width=16, fig.height=12}

pState_1 <- ggplot(aes(x=reorder(BorrowerState, BorrowerState, 
                                 function(x) { length(x)})), 
                   data = new_pl) +
  geom_bar() +
  geom_text(
    stat = 'count',
    aes(label = ..count..),
    position = position_dodge(0.9),
    hjust = -0.25
  ) +
  ylim(0, 11000) +
  ggtitle('# Applications by BorrowerState') +
  coord_flip()

pState_2 <- ggplot(aes(x=reorder(BorrowerState, BorrowerState, 
                                 function(x) { length(x)}), 
                       y=LoanOriginalAmount), 
                   data=new_pl) +
  geom_boxplot(width=0.5, position=position_dodge(width=0.25)) +
  ggtitle('LoanOriginalAmount by BorrowerState') +
  coord_flip()

grid.arrange(pState_1, pState_2, ncol=2)

```

From the summary in terms of a number of applications in each state, 'Wyoming' 
(WY) has the lowest number of applications which is 119 and 'California' (CA) 
has the highest number of applications which is 10,479.

The median number of applications is 1161.5 and average number is 1722.4.

```{r echo=FALSE}

grp_BorrowerState <- new_pl %>% 
  group_by(BorrowerState) %>% 
  summarise(Count = n()) %>%
  arrange(desc(Count))

summary(grp_BorrowerState$Count)
```

The boxplot below shows that there are four states where a number of 
applications far exceeds the majority (outliers). 

When combining information from the bar chart above and the boxplot below we can
identify that those four states are California (CA), New York (NY), Texas (TX),
and Florida (FL). Also notice that a number of applications in CA is also far 
more than the other three states.

```{r echo=FALSE}

ggplot(aes(x='', y=Count), data= grp_BorrowerState) +
  geom_boxplot() +
  xlab('# applications in each state')
```


The maps below make it clear that a number of listings is higher in the major 
cities.

* This first map clearly shows that there are three states loans are 
not available; Maine, Iowa, and North Dakota
(Complaint#5 in https://www.lendingmemo.com/prosper-loan-complaints/)
* The second map shows that the big cities have a higher number of applications


```{r echo=FALSE, fig.width=10}

grp_BorrowerState_map <- new_pl %>% 
  mutate(statename = state.name[match(BorrowerState,state.abb)]) %>%
  group_by(statename) %>% 
  summarise(Count = n()) %>% 
  mutate(long = state.center$x[match(statename, state.name)], 
         lat = state.center$y[match(statename, state.name)])

states_map <- map_data("state")
ggplot(grp_BorrowerState_map, aes(map_id = tolower(statename))) +
    geom_map(aes(fill = Count), map = states_map) +
    expand_limits(x = states_map$long, y = states_map$lat)
  
```

```{r echo=FALSE, fig.width=10}

ggplot() +
  geom_path(data = states_map, 
            aes(x = long, y = lat, group = group)) +
  geom_point(data = subset(grp_BorrowerState_map, !is.na(statename)), 
             aes(x = long, y = lat, size = Count), 
             shape = 21, colour = "#000000", fill = "#40b8d0") 

```

__Term__

The histogram below shows that most applications have the 36-Term.

```{r echo=FALSE}

grp_Term <- new_pl %>% 
  group_by(Term) %>% 
  summarise(Count = n())


ggplot(aes(x=Term, y=Count), data=grp_Term) +
  geom_bar(stat='identity') +
  geom_text(
    aes(label = Count),
    position = position_dodge(0.9),
    vjust = -0.5
  )

```

__IncomeRange__

Majority of the applications have the IncomeRange between $1-$50,000. There are 
also applications with IncomeRange = 0. Did they get a loan?

```{r echo=FALSE}

grp_IncomeRange <- new_pl %>% 
  group_by(IncomeRange) %>% 
  summarise(Count = n())


ggplot(aes(x=IncomeRange, y=Count), data=grp_IncomeRange) +
  geom_bar(stat='identity')


```

__DebtToIncomeRatio__

From the histogram and boxplot, most of the applications have the 
DebtToIncomeRatio less than 1. 

However, there are also outliers where the DebtToIncomeRatio > 1 and up to 
DebtToIncomeRatio = 10.01. What is a status of those applications?

```{r echo=FALSE}

p1 <- ggplot(aes(x=DebtToIncomeRatio), 
             data=subset(new_pl, is.finite(DebtToIncomeRatio))) +
  geom_histogram(binwidth=0.05)

p2 <- ggplot(aes(x='', y=DebtToIncomeRatio), 
             data= subset(new_pl, is.finite(DebtToIncomeRatio))) +
  geom_boxplot() +
  coord_flip()

grid.newpage()
grid.draw(rbind(ggplotGrob(p1), ggplotGrob(p2), size = "last"))

```



```{r echo=FALSE}

summary(subset(new_pl, is.finite(DebtToIncomeRatio))$DebtToIncomeRatio)
```

The frequency polygon below shows that most applications have the DebtToIncomeRatio 
around 0.2.

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(aes(x=DebtToIncomeRatio), data=subset(new_pl,  DebtToIncomeRatio < 2)) +
  geom_freqpoly(binwidth=0.01) +
  scale_x_continuous(limits = c(0, 1))
```

__ListingCategory__

Most applications apply the Proper loan for 'Debt Consolidation' with the median 
LoanOriginalAmout around $10,000

```{r echo=FALSE, , fig.width=10}

p1 <- ggplot(aes(x=reorder(ListingCategory, ListingCategory, 
                           function(x) length(x))), 
             data=new_pl) +
  geom_bar() +
  ggtitle('# applications by ListingCategory') +
  coord_flip()

p2 <- ggplot(aes(x=reorder(ListingCategory, ListingCategory, 
                           function(x) { length(x)}), 
                 y=LoanOriginalAmount), 
             data=new_pl) +
  geom_boxplot(width=0.5, position=position_dodge(width=0.25)) +
  ggtitle('LoanOriginalAmount by ListingCategory') +
  coord_flip()

grid.arrange(p1, p2, ncol=2)
```


__CreditScoreRange__

The median of the CreditScoreRangeLower is 700 and the median of the 
CreditScoreRangeUpper is 719.

```{r echo=FALSE, fig.height=8}

p1 <- ggplot(aes(x=CreditScoreRangeLower), data=new_pl) +
  geom_histogram(binwidth=5) +
  xlim(600, 900)

p2 <- ggplot(aes(x='', y=CreditScoreRangeLower), data= new_pl) +
  geom_boxplot() +
  ylim(600, 900) +
  coord_flip()

pUpper1 <- ggplot(aes(x=CreditScoreRangeUpper), data=new_pl) +
  geom_histogram(binwidth=5)  +
  xlim(600, 900)

pUpper2 <- ggplot(aes(x='', y=CreditScoreRangeUpper), data= new_pl) +
  geom_boxplot()  +
  ylim(600, 900) +
  coord_flip()

grid.newpage()
grid.draw(rbind(ggplotGrob(p1), 
                ggplotGrob(p2), 
                ggplotGrob(pUpper1),
                ggplotGrob(pUpper2),
                size = "first"))

```


```{r echo=FALSE}
print('*** Summary of CreditScoreRangeLower ***')
summary(new_pl$CreditScoreRangeLower)
```


```{r echo=FALSE}

print('*** Summary of CreditScoreRangeUpper ***')
summary(new_pl$CreditScoreRangeUpper)
```


# Univariate Analysis

### What is the structure of your dataset?

There are approximately 114,000 records and 81 variables. After analyzing the 
ListingCreatingDate variable, We found that data with various CreditGrade were
not created after 2009, and only data with the CreditGrade 'AA' was created 
after 2010. Hence, I only filtered data with the ListingCreatingDate is greater
than or equal to 2010 for this exploration.

With the new dataset, there are approximately 83,000 records with 82 variables 
where an additional factor variable 'ListingCategory' was created to store a 
ListingCategory name of each record


### What is/are the main feature(s) of interest in your dataset?

I am paricularly interested in finding variables that are best for describing 
relationship with the BorrowerRate variable.


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I think the following variables will correlate to the BorrowerRate; 
Occupation, EmploymentStatus, BorrowerState, Term, IncomeRange, 
DebtToIncomeRatio, ListingCategory, CreditScoreRangeLower, and 
CreditScoreRangeUpper.
 

### Did you create any new variables from existing variables in the dataset?

As mentioned in the previous section, I created a new variable 'ListingCategory'
to store a text of each category to accomodate this analysis.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I found an unsual trend in the charts that show a number of listings based 
on the ListingCreatingDate and found that there are no records with various 
CreditGrade after 2010, and only records with the CreditGrade 'AA' have the
ListingCreatingDate greater than or equal to 2010 so I decided to only 
analyze records where the ListingCreatingDate >= 2010.

Also, I updated levels of the variables 'CreditGrade' and 'IncomeRange' to ease 
a process of creating visualizations for those variables so that an appropriate
level's order is shown in the plots.


# Bivariate Plots Section


__Scatterplot Matrix__

Let's start off by creating a scatterplot matric with the variables below:

BorrowerRate, LoanOriginalAmount, StatedMonthlyIncome, DebtToIncomeRatio, CreditScoreRangeLower, CreditScoreRangeUpper, Term, EmploymentStatus

Note that we did not choose the 'Occupation', 'BorrowerState' and 
'ListingCategory' in this plot because they have a lot of levels

The correlation matrix shows that the 'LoanOriginalAmount', 
'CreditScoreRangeLower', and 'CreditScoreRangeUpper' variables have a strong 
correlation with the 'BorrowerRate' variable


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=12}

new_pl_bi <-  select(new_pl,
                     BorrowerRate, LoanOriginalAmount, StatedMonthlyIncome,
                     DebtToIncomeRatio, CreditScoreRangeLower,
                     CreditScoreRangeUpper, Term, EmploymentStatus)

ggcorr(new_pl_bi, label = TRUE)

```

__BorrowerRate vs LoanOriginalAmount__

Let's take a look at a scatter plot of BorrowerRate (y) vs LoanOriginalAmount(x)

```{r echo=FALSE, message=FALSE}

ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate), data=new_pl_bi) +
  geom_point(alpha=1/50, position = 'jitter') +
  geom_smooth()
```

The correlation coefficient between BorrowerRate and  LoanOriginalAmount is 
around -0.42 which shows a moderate correlation between those two variables

```{r echo=FALSE}

with(new_pl_bi, cor.test(LoanOriginalAmount, BorrowerRate))

```

__BorrowerRate vs CreditScoreRange__

Let's take a look at a scatter plot of BorrowerRate (y) vs 
CreditScoreRangeLower(x) and a scatter plot of BorrowerRate (y) vs 
CreditScoreRangeUpper(x)

The smoothed conditional mean line shows that the BorrowerRate is decreased when
the CreditScoreRangeLower or CreditScoreRangeUpper is increased

```{r echo=FALSE, message=FALSE}

p1 <- ggplot(aes(x=CreditScoreRangeLower, y=BorrowerRate), data=new_pl_bi) +
  geom_point(alpha=1/25, position = 'jitter', color='orange') +
  geom_smooth() +
  ggtitle('Scatter plot of BorrowerRate (y) vs CreditScoreRangeLower(x)')

p2 <- ggplot(aes(x=CreditScoreRangeUpper, y=BorrowerRate), data=new_pl_bi) +
  geom_point(alpha=1/25, position = 'jitter', color='orange') +
  geom_smooth() +
  ggtitle('Scatter plot of BorrowerRate (y) vs CreditScoreRangeUpper(x)')

grid.arrange(p1, p2, ncol=1)

```

The correlation coefficient between BorrowerRate and CreditScoreRangeLower is 
around -0.5 which shows a moderate correlation between those two variables

```{r echo=FALSE}

with(new_pl_bi, cor.test(CreditScoreRangeLower, BorrowerRate))

```

The correlation coefficient between BorrowerRate CreditScoreRangeUpper is 
around -0.5 which shows a moderate correlation between those two variables

```{r echo=FALSE}

with(new_pl_bi, cor.test(CreditScoreRangeUpper, BorrowerRate))

```


__BorrowerRate vs BorrowerState__

The boxplot below confirms that the BorrowerState does not have impact on the 
BorrowerRate since the median value of the BorrowerRate of all states is around 
0.2 and the boxplots are closely aligned.

```{r echo=FALSE, fig.width=12}

ggplot(aes(x=BorrowerState , y=BorrowerRate), data=new_pl) +
  geom_boxplot() +
  stat_summary(fun.y=median, geom="smooth", aes(group=1), lwd=1)

```


__BorrowerRate vs Occupation__

The boxplot below also shows that Occuputation does not correlate to the 
BorrowerRate since the median BorrowerRate is around 0.2 excepts the Occupations 
with a small number of applications that shows lower BorrowerRate, (for example,
Judge) 

```{r echo=FALSE, fig.width=16, fig.height=12}

p1 <- ggplot(aes(x=reorder(Occupation, Occupation, 
                           function(x) length(x))), 
             data = new_pl) +
  geom_bar() +
  ggtitle('# Applications by Occupation') +
  coord_flip()


p2 <- ggplot(aes(x = reorder(Occupation, Occupation, 
                           function(x) { length(x)}), 
                 y = BorrowerRate), 
             data = new_pl) +
  geom_boxplot(width=0.5, position=position_dodge(width=0.25)) +
  ggtitle('BorrowerRate by Occupation') +
  stat_summary(fun.y=median, geom="smooth", aes(group=1), lwd=1) +
  coord_flip()

grid.arrange(p1, p2, ncol=2)
```

__BorrowerRate vs ListingCategory__

The boxplot below also shows that ListingCategory does not correlate to the 
BorrowerRate since the median BorrowerRate is around 0.2.

```{r echo=FALSE, fig.width=16, fig.height=12}


p1 <- ggplot(aes(x=reorder(ListingCategory, ListingCategory, 
                           function(x) length(x))), 
             data = new_pl) +
  geom_bar() +
  ggtitle('# applications by ListingCategory') +
  coord_flip()

p2 <- ggplot(aes(x = reorder(ListingCategory, ListingCategory, 
                           function(x) { length(x)}), 
                 y = BorrowerRate), 
             data = new_pl) +
  geom_boxplot(width=0.5, position=position_dodge(width=0.25)) +
  ggtitle('BorrowerRate by ListingCategory') +
  stat_summary(fun.y=median, geom="smooth", aes(group=1), lwd=1) +
  coord_flip()

grid.arrange(p1, p2, ncol=2)
```

__BorrowerRate vs StatedMonthlyIncome__

Let's take a look at a scatter plot of BorrowerRate (y) vs StatedMonthlyIncome(x).

This scatter plot shows multiple outliers that is difficult to see a relation 
between the BorrowerRate and StatedMonthlyIncome.

```{r echo=FALSE}

ggplot(aes(x=StatedMonthlyIncome, y=BorrowerRate), data=new_pl_bi) +
  geom_point()

```

So what is a summary of StatedMonthlyIncome?

```{r echo=FALSE}

summary(new_pl_bi$StatedMonthlyIncome)
```

Let's create a scatter plot of BorrowerRate (y) vs StatedMonthlyIncome(x) where
StatedMonthlyIncome is in the 95% percentile.

It looks like the BorrowerRate is slightly decreased when the 
StatedMonthlyIncome is increased.

```{r echo=FALSE, message=FALSE}

d_Income_95 <- subset(new_pl_bi, 
                      StatedMonthlyIncome <= quantile(StatedMonthlyIncome, 
                                                      0.95) )

ggplot(aes(x=StatedMonthlyIncome, y=BorrowerRate), 
       data=subset(d_Income_95, StatedMonthlyIncome > 0)) +
  geom_point(alpha=1/25, position='jitter', color='orange') +
  geom_smooth()

```

The correlation coefficient between BorrowerRate and  StatedMonthlyIncome is 
around -0.25 which shows a weak correlation between those two variables

```{r echo=FALSE}

with(d_Income_95, cor.test(StatedMonthlyIncome, BorrowerRate))

```

__BorrowerRate vs DebtToIncomeRatio__
Let's take a look at a scatter plot of BorrowerRate (y) vs DebtToIncomeRatio(x).

This scatter plot shows multiple outliers that is difficult to see a relation 
between the BorrowerRate and DebtToIncomeRatio

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x=DebtToIncomeRatio, y=BorrowerRate), data=new_pl_bi) +
  geom_point()

```


What is a summary of DebtToIncomeRatio?

```{r echo=FALSE}

summary(new_pl_bi$DebtToIncomeRatio)
```

Let's create a scatter plot of BorrowerRate (y) vs DebtToIncomeRatio(x) where
DebtToIncomeRatio is in the 99% percentile.

It looks like the BorrowerRate is increased when the DebtToIncomeRatio is 
increased.

```{r echo=FALSE, message=FALSE}

d_DebtToIncomeRatio_99 <- subset(new_pl_bi, 
                      DebtToIncomeRatio <= quantile(DebtToIncomeRatio, 0.99, 
                                                    na.rm = TRUE) )

ggplot(aes(x=DebtToIncomeRatio, y=BorrowerRate), 
       data=subset(d_DebtToIncomeRatio_99, DebtToIncomeRatio > 0)) +
  geom_point(alpha=1/25, position='jitter', color='orange') +
  geom_smooth()

```

The correlation coefficient between BorrowerRate and  DebtToIncomeRatio is 
around 0.18 which shows a weak correlation between those two variables

```{r echo=FALSE}

with(d_DebtToIncomeRatio_99, cor.test(DebtToIncomeRatio, BorrowerRate))

```

__LoanOriginalAmount vs StatedMonthlyIncome__

Let's take a look at the relationship between LoanOriginalAmount and 
StatedMonthlyIncome.

The scatter plot of LoanOriginalAmount(y) and 95 percentile of 
StatedMonthlyIncome (x) shows that the LoanOriginalAmount is increased when the 
StatedMonthlyIndome is increased.

```{r echo=FALSE, message=FALSE}

d_StatedMonthlyIncome_95 <- subset(new_pl, 
                      StatedMonthlyIncome <= quantile(StatedMonthlyIncome, 0.95) )
ggplot(aes(x=StatedMonthlyIncome, y=LoanOriginalAmount), 
       data=d_StatedMonthlyIncome_95) + 
  geom_point(alpha=1/50, position='jitter', color = 'orange') +
  geom_smooth()
  
```

The correlation coefficient between LoanOriginalAmount and 95 percentile of 
StatedMonthlyIncome is 0.41 which means that these two variables have a moderate
correlation.

```{r echo=FALSE}

with(d_StatedMonthlyIncome_95, cor.test(StatedMonthlyIncome, LoanOriginalAmount))

```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

There are a few features, which are LoanOriginalAmount, CreditScoreRangeLower, 
CreditScoreRangeLower, that have a moderate correlation with the BorrowerRate.

Additionally, there are a few, which are StatedMonthlyIncome and 
DebtToIncomeRatio, that shows a weak correlation with the BorrowerRate after 
removing the outlier.

However, other features, such as BorrowerState, Occupation, ListingCategory, do 
not seem to have any strong relationship with the BorrowerRate

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

The StatedMonthlyIncome and LoanOriginalAmount variables have a moderate
correlation where the LoanOriginalAmount is increased when the 
StatedMonthlyIncome is increased

### What was the strongest relationship you found?

The CreditScoreRangeLower/Update have the strongest relationship with the 
BorrowerRate where their correlation coefficient is around -0.5

# Multivariate Plots Section


__BorrowerRate, LoanOriginalAmount | Term __

Let's take a look at a scatter plot of BorrowerRate (y) vs LoanOriginalAmount(x)
and use a different color for each Term.

It looks like most smaller LoanOriginalAmout has 36 Term but it is hard to tell
because of overplotting. Let's use a facet_wrap instead.

```{r echo=FALSE}

ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate), data=new_pl) +
  geom_point(aes(colour = Term)) 

```

The plots below show that most of the applications have lower LoanOriginalAmount
for all Terms.
Also, it looks like the LoanOriginalAmount is usually about x$5000 
(darker points) as we see in the plots

```{r echo=FALSE, , message=FALSE}

ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate), data=new_pl) +
  geom_point(alpha=1/50, position = 'jitter') +
  facet_wrap(~Term, ncol=1)

```

The conditional mean chart for different Term shows that the BorrowerRate of the
12-Term is the lowest regardless of the LoanOriginalAmount, follow by the 
36-Term and 60-Term.

```{r echo=FALSE, message=FALSE}
ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate), data=new_pl) +
  geom_smooth(aes(color=Term))

```

__BorrowerRate, LoanOriginalAmount | Occupation, BorrowerState, and ListingCategory__

The facet plots below show that Occupation, BorrowerState, and ListingCategory
does not have a strong relationship with the BorrowRate and LoanOriginal amount
since all conditional mean plots show a similar Trend.

Additionally, most applications with the LoanOriginalAmout less than $10,000 
have the 36-Term with various BorrowerState regardless of Occupation, 
BorrowerState, and ListingCategory.

```{r echo=FALSE, fig.width=16, fig.height=16, message=FALSE, warning=FALSE}


ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate), data=new_pl) +
  geom_point(alpha=1/20, position = 'jitter') +
  geom_smooth() +
  facet_wrap(~Occupation, ncol=8)


```

```{r echo=FALSE, fig.width=16, fig.height=16, message=FALSE}


ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate, color=Term), data=new_pl) +
  geom_point(alpha=1/2) +
  facet_wrap(~Occupation, ncol=8)


```

```{r echo=FALSE, fig.width=16, fig.height=16, message=FALSE}


ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate), data=new_pl) +
  geom_point(alpha=1/20, position = 'jitter') +
  geom_smooth() +
  facet_wrap(~BorrowerState, ncol=8)


```

```{r echo=FALSE, fig.width=16, fig.height=16, message=FALSE}


ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate, color=Term), data=new_pl) +
  geom_point(alpha=1/2) +
  facet_wrap(~BorrowerState, ncol=8)


```


```{r echo=FALSE, fig.width=16, fig.height=16, message=FALSE}


ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate), data=new_pl) +
  geom_point(alpha=1/20, position = 'jitter') +
  geom_smooth() +
  facet_wrap(~ListingCategory, ncol=5)


```


```{r echo=FALSE, fig.width=16, fig.height=16, message=FALSE}


ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate, color=Term), data=new_pl) +
  geom_point(alpha=1/2) +
  facet_wrap(~ListingCategory, ncol=5)


```

__BorrowerRate, LoanOriginalAmount | EmploymentStatus __

The chart below shows that applications with the EmploymentStatus 'Not employed' 
usually gets a higher BorrowerRate regardless of the LoanOriginalAmount where 
the BorrowRate for other EmploymentStatus is in the same range

```{r echo=FALSE, message=FALSE}

ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate), data=new_pl) +
  geom_smooth(aes(color=EmploymentStatus))
```

__BorrowerRate, LoanOriginalAmount | IncomeRange __

The chart below shows that applications with the IncomeRange 'Not displayed' gets
a higher BorrowerRate in general.

However, what surprise me is that the applications with the highest IncomeRange
($100,000) get a much higher BorrowerRate that other IncomeRange and 
applications with the lowest IncomeRange ($0) get the lowest BorrowerRate

```{r echo=FALSE, message=FALSE}
ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate), data=new_pl) +
  geom_smooth(aes(color=IncomeRange))

```

__BorrowerRate, LoanOriginalAmount | IncomeRange __

As expected, applications with a lower CreditScoreRange (Lower and Upper) get 
the Lower BorrowerRate.

Interestingly, applications with a very high CreditScoreRange (Lower and Upper) 
get the BorrowerRate around 0.1 which is much lower than other CreditScoreRanges.

```{r echo=FALSE, message=FALSE}

new_pl$CreditScoreRangeLower.bucket <- cut(new_pl$CreditScoreRangeLower, 
                                           breaks = seq(550, 900, by = 50))

# Need to create a custom color palette to make the category name consistent
myColors <- brewer.pal(7,"Set1")
names(myColors) <- levels(new_pl$CreditScoreRangeLower.bucket)

p1 <- ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate), data=new_pl) +
  geom_smooth(aes(color=CreditScoreRangeLower.bucket)) +
  scale_colour_manual(name = "CreditScoreRangeLower.bucket",values = myColors)


new_pl$CreditScoreRangeUpper.bucket <- cut(new_pl$CreditScoreRangeUpper, 
                                           breaks = seq(550, 900, by = 50))

p2 <- ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate), data=new_pl) +
  geom_smooth(aes(color=CreditScoreRangeUpper.bucket)) +
  scale_colour_manual(name = "CreditScoreRangeUpper.bucket",values = myColors)


grid.newpage()
grid.draw(rbind(ggplotGrob(p1), 
                ggplotGrob(p2),
                size = "first"))

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

There are a few features that provide better insights on the BorrowerRate when 
observing them with LoanOriginalAmount.

Those features are 
* Term: The BorrowerRate is less with a smaller term and the BorrowRate will 
keep decreasing as the LoanOriginalAmount is increased
* EmploymentStatus: The BorrowerRate is the highest when the EmploymentStatus is
'Not employed'
* CreditScoreRange: The BorrowerRate is higher when the CreditScore is less

### Were there any interesting or surprising interactions between features?

I am surprised to see that applications with the highest IncomeRange get a much
higher BorrowerRate than applications with a very small IncomeRange (e.g. $0).

I originally thought that the IncomeRange should not have any relationship with
the BorrowerRate but it does not seem to be the case.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I did not create any models with this dataset.
------

# Final Plots and Summary

### Plot One

```{r echo=FALSE, Plot_One}

ggplot(aes(x=date_by_month, y=n), data=grpCreationDate_CreditGrade) + 
  geom_line() + 
  xlab('Listing creation on (in months)') +
  ylab('Number of listings') +
  scale_x_date( date_labels = "%Y",date_breaks = "2 year") + 
  facet_wrap(~CCGrade, ncol=3) +
  ggtitle('THE NUMBER OF LISTINGS OF EACH CREDIT GRADE')
```

### Description One

I chose this plot for the Plot One section because it shows that majority of 
listings in this dataset has the CreditGrade 'AA' and they were created after 
2009 as opposed to other CreditGrades which have a much smaller number of 
listings comparing to the CreditGrade 'AA' and there are no listings with those
CreditGrades after 2009.

This finding help me make a decision to only analyze data of the listings that 
were created after 2009 in this analysis.

### Plot Two

```{r echo=FALSE, message=FALSE, Plot_Two}

p1 <- ggplot(aes(x=CreditScoreRangeLower, y=BorrowerRate), 
             data=new_pl_bi) +
  geom_point(alpha=1/40, position = 'jitter', color='orange') +
  geom_smooth() +
  xlab('Lower bound of the credit score range') +
  ylab("Borrower's interest rate") +
  coord_cartesian(c(575, 900)) +
  ggtitle('BorrowerRate vs CreditScoreRangeLower')

p2 <- ggplot(aes(x=CreditScoreRangeUpper, y=BorrowerRate), 
             data=new_pl_bi) +
  geom_point(alpha=1/40, position = 'jitter', color='orange') +
  geom_smooth() +
  xlab('Upper bound of the credit score range') +
  ylab("Borrower's interest rate") +
  coord_cartesian(c(575, 900)) +
  ggtitle('BorrowerRate vs CreditScoreRangeUpper')

grid.newpage()
grid.draw(rbind(ggplotGrob(p1), 
                ggplotGrob(p2),
                size = "first"))

```

### Description Two

This chart clearly shows a strong relationship between the BorrowerRate and 
CreditScoreRange (Lower/Upper) where the conditional mean of the BorrowerRate is
decreased as the credit score range increased.

### Plot Three

```{r echo=FALSE, message=FALSE, Plot_Three, fig.width=8}

ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate), data=new_pl) +
  geom_smooth(aes(color=Term)) +
  xlab('Original amount of the loan (dollars)') +
  ylab("Borrower's interest rate") + 
  labs(color='Length of the loan') +
  ggtitle("Change of the Borrower's interest rate by the original amount and length of the loan")
```

### Description Three

This chart shows that the borrower's interest rate is much lower for the 12-Term 
in average, followed by 36 and 60 Term, and as the original amound of the loan
increased, the Borrower's interest rate is decreased.

------

# Reflection

This dataset contains a large number of variables and it was challenging to come
up with a question and choose main and other features to analyze because there 
are so  many of them.

Additionally, since I do not have knowledge about Prosper Loan before working on 
this project, I had a difficulty of the terms and technical details explained in
the variable definition and this also impacts on the previous problem where I 
spent a lot of time coming up with a question and choose variables that I think 
it might with the analysis and, as I went through the exercise, many variables 
that I chose did not seem to have a relationship with the main feature.

After spending time understanding the dataset and getting a basic understanding 
of the ProsperLoan, I was able to identify states which does not allow Prosper 
Loan by inspecting the geom_map and found the backup information after this 
analysis.

In the Univariate analysis section, I also discovered that there are a handful of 
listing data for various CreditGrade that was stopped creating after 2009 and 
only the listing data with the CreditGrade 'AA' was created after 2009 and this 
significantly help me make a decision to only analyze the listing data created 
after 2009 which is the majority of the population.

Most of the variables that I chose in the Bivariate analysis does not have a 
strong relationship with the main feature (BorrowerRate) which I think that the 
biggest problem for this is because of a lack of domain knowledge as I mentioned 
earlier.

I found the Multivariate Analysis is quite interesting since it help giving 
insights for the interactions between BorrowerRate, LoanOriginalAmount and other
variables. Especially, I was surprised to see that the listing with the highest
IncomeRange get a much higer BorrowerRate on average comparing to the listing 
with a very low income (e.g. $0)

This dataset has rich features that can be used in the future analysis. I have 
tidied a few variables to help with this analysis, for example, re-arrange 
factor levels of the IncomeRange and CreditGrade variables and create a new 
variable to store a ListingCategory name (the original dataset only contains the
Category index). There are many more variables in this dataset need to be tidied 
to accomodate other Exploratory Data Analysis in the future.